---
title: "02_biomarker imputation"
author: "Jagadeesh Puvvula"
date: "2026-03-09"
output: pdf_document
---

#PBDE exposures
```{r}
pbde_df <- haven::read_sas("~/Documents/data/pbde.sas7bdat") |> clean_names() |>
  mutate(across(any_of(c("subject_id", "visit", "analyte_code")), factor),
                across(contains("flg_lod"), factor)) |>
  rename(flg_lod = flg_lod_lp,
         result = result_lp) |>
  select(c(subject_id, visit, analyte_code, result, flg_lod, analyte_lod_lp)) |>
  filter(visit %in% c("16W", "26W"))
```

#pthalate, OPE, cotinine, and covariates
```{r}
phthalate_df <- haven::read_sas("~/Documents/data/phthbpa.sas7bdat") |> clean_names() |>
  mutate(across(any_of(c("subject_id", "visit", "analyte_code")), factor),
                across(contains("flg_lod"), factor)) |>
  select(c(subject_id, visit, analyte_code, result, flg_lod, analyte_lod))|>
  filter(visit %in% c("16W", "26W"),
         !analyte_code %in% c("MCNP", "MCOP")) 

ope_df <- haven::read_sas("~/Documents/data/fr.sas7bdat") |> clean_names() |>
  mutate(across(any_of(c("subject_id", "visit", "analyte_code")), factor),
                across(contains("flg_lod"), factor)) |>
  select(c(subject_id, visit, analyte_code, result, flg_lod, analyte_lod))|>
  filter(visit %in% c("16W", "26W"))

cot_df <- haven::read_sas("~/Documents/data/ct.sas7bdat") |> clean_names() |>
  mutate(result = if_else(is.na(result), result_machine, result)) |>
  select(c(subject_id, visit, analyte_code, result, flg_lod, analyte_lod)) |>
  mutate(across(any_of(c("subject_id", "visit", "analyte_code")), factor),
                across(contains("flg_lod"), factor)) |>
  filter(visit %in% c("16W", "26W"))

cov_df<- haven::read_sas("~/Documents/data/covariates.sas7bdat") |> clean_names() |>
  select(-c(race_cd, ethnicity_cd, mom_race_cd, mom_race, mom_ethnicity_cd, hh_income,
         mom_ethnicity, hh_income_cd, mari_st_cd, gest_diabetes, gest_hypertension,
         mom_edu_cd, preeclampsia, gest_cardio_comp, parity_cat, gest_age_method, fish_consumption_cd)) 
```

# %% count of biomarker measures above and below LOD
```{r}
counts_all <- bind_rows(
  ope_df        |> mutate(dataset = "OPEs"),
  phthalate_df  |> mutate(dataset = "Phthalate"),
  pbde_df       |> mutate(dataset = "PBDEs")
) |>
  filter(subject_id %in% basc_out$subject_id) |>
  group_by(dataset, analyte_code, visit) |>
  summarise(
    n = n(),
    pct_lod = mean(flg_lod == 1, na.rm = TRUE) * 100,
    .groups = "drop"
  ) |>
  mutate(n_label = sprintf("%d (%.1f%%)", n, pct_lod)) |>
  select(dataset, analyte_code, visit, n_label) |>
  pivot_wider(
    names_from = visit,
    values_from = n_label,
    values_fill = "0 (0%)"
  )
```

#covariate imputation
```{r}
cov_df <- read_sas("~/Documents/data/covariates.sas7bdat") |>
  clean_names() |>
  select(c(1,2,4,6,12,15,17,34,54,55,57,64))|>
  mutate(
    race = case_when(
      race == "White or Caucasian" ~ "White",
      race == "Black or African American" ~ "Black",
      race %in% c("Asian, Pacific Islander, or Native Hawaiian",
                  "Other/Multiracial") ~ "Other",
      TRUE ~ NA_character_),
    ethnicity = case_when(
      ethnicity == "Hispanic" ~ "Hispanic",
      ethnicity == "Not Hispanic" ~ "Non-Hispanic",
      TRUE ~ NA_character_),
    income_group = case_when(
      hh_income %in% c(
        "Under $5,000","Between $5,000 and $10,000","Between $10,000 and $15,000",
        "Between $15,000 and $20,000", "Between $20,000 and $25,000",
        "Between $25,000 and $30,000","Between $30,000 and $40,000") ~ "Low",
      hh_income %in% c(
        "Between $40,000 and $50,000","Between $50,000 and $60,000",
        "Between $60,000 and $70,000","Between $70,000 and $80,000") ~ "Middle",
      hh_income %in% c(
        "Between $80,000 and $90,000","Between $90,000 and $100,000",
        "Between $100,000 and $110,000","Between $110,000 and $120,000",
        "Between $120,000 and $130,000","Between $130,000 and $140,000",
        "Between $140,000 and $150,000","Over $150,000") ~ "High"),
    mom_edu = case_when(
      mom_edu %in% c(
      "8th grade or less","Some high School (Grades 9, 10, 11, 12)",
      "High school diploma (completed grade 12)",
      "GED","Technical or trade school") ~ "L1",
      mom_edu == "Some college or 2 year degree" ~ "L2",
      mom_edu == "Bachelor's degree" ~ "L3",
      mom_edu == "Graduate or professional school" ~ "L4",
      TRUE ~ NA_character_),
    mari_st = case_when(
      mari_st == "Married, living together" ~ "Marr_liv_toget",
      mari_st %in% c("Married, living apart",
                     "Not married, but living with someone",
                     "Not married, living alone") ~ "Other",
      TRUE ~ NA_character_),
    fish_baseline = case_when(
      fish_consumption_cd %in% c("1","2", "3") ~ "L1",
      fish_consumption_cd %in% c("4") ~ "L2",
      fish_consumption_cd %in% c("5") ~ "L3",
      fish_consumption_cd %in% c("6") ~ "L4",
      TRUE ~ NA_character_),
  ) |>
  mutate(across(where(is.character), as.factor)) |>
  select(-c(hh_income, ethnicity, fish_consumption_cd))

#impute missing covariates
categorical_vars <- c("sex", "race", "mari_st", "mom_edu",
                      "income_group", "fish_baseline")
continuous_vars <- c("home_score_total", "mom_age_deliv", "gest_age",
                     "parity_n")

# Run imputation
result <- impute_with_superlearner(cov_df,
                                   categorical_vars = categorical_vars, 
                                   max_predictors = 10,
                                   verbose = TRUE)

save(result, file = "~/Documents/HOME_DNAm/covariate_imputation.RData")
write_csv(result$imputed_data, "~/Documents/HOME_DNAm/covariate_imputation.csv")
```

#load covariate data
#use imputed covariate data for imputing biomarker data
```{r}
cov<- read_csv("~/Documents/k99_home/data/covariate_imputation.csv") |>
  filter(subject_id %in% basc_out$subject_id) |>
  mutate(subject_id = as.factor(subject_id)) 
```


# %% correlations between exposures and covariates
```{r}
biomarker_df <- bind_rows(
  phthalate_df |> mutate(exposure = "phthalates"),
  ope_df |> mutate(exposure = "organophosphate esters"),
  cot_df |> mutate(exposure = "cotinine")
) |>
  mutate(analyte_visit = paste(analyte_code, visit, sep = "_")) |>
  select(-c(analyte_code, visit))  |>
  filter(subject_id %in% basc_out$subject_id)

cov_pred_df<- cov_df |>
  select(c(1:7, 24, 48)) |>
  mutate(subject_id = as.factor(subject_id))

lod_values <- biomarker_df |>
  mutate(analyte = sub("_.*$", "", analyte_visit)) |>
  select(analyte, analyte_lod) |>
  distinct() 

lod_map <- setNames(lod_values$analyte_lod, lod_values$analyte)
```

#use values in the variable result;
#dont use result_original since it has NAs for values below LOD & result_imputed which has values on log scale (gives negative values)
```{r}
result <- truncated_single_imputation_sl(
  biomarker_df      = biomarker_df,
  cov_pred_df       = cov,
  lower_bounds      = 0.00001, 
  maxit_truncated   = 100,
  sl_library        = c("SL.glm", "SL.glmnet", "SL.randomForest",
                        "SL.xgboost", "SL.mean"),
  log_transform     = FALSE,
  adaptive_weight   = TRUE,
  track_convergence = TRUE,
  verbose           = TRUE
  )
```

```{r}
imputed     <- get_imputed_data(result) |> select(c(1:6))
metrics     <- get_quality_metrics(result, NULL)
sl_only     <- get_quality_metrics(result, "success")
fallbacks   <- metrics %>% filter(imputation_method == "conventional")

save(imputed, metrics, sl_only, fallbacks, result,
     file = "~/Documents/k99_home/data/exposure_impu/exposure_imputation.RData")

write_csv(imputed, "~/Documents/k99_home/data/exposure_impu/imputed_exposures.csv")
```



