---
title: "Preprocessing DNA methylation data EPIC array"
author: "Puvvula"
date: "2025-12-24"
output: pdf_document
---

```{r}
library(pacman)
pacman::p_load(tidyverse, janitor, haven)
```

#read metadata
```{r}
#low quality on original and repeat: 79570, 47315, 48087, 48087, 50695, 43374, 87028, 16716, 32700, 46349, 66865, 95834, 49157, 87028, 43374, 46349, 66865, 95834, 48087, 50695, 47315, 79570, 32700, 16716, 
#low quality not repeated: 46864, 21266, 24648, 

#low quality on original but ok on repeat: 48031, 33204, 44919
#57128, 86041, 57128, 86041,  not sure about "Two sections red; included in repeat"

meta <- readxl::read_xlsx(
  "~/Documents/k99_home/DNA_Methylation_FileList_Birth_Cord_DeID.xlsx"
) |>
  janitor::clean_names() |>
  dplyr::filter(
    !subject_id %in% c(
      79570, 47315, 48087, 50695, 43374, 87028, 16716, 32700,
      46349, 66865, 95834, 49157, 46864, 21266, 24648),
    !comments %in% c("Low quality scanned data",
                     "Include in repeat",
                     "Two sections red; include in repeat")
  )

#check that each subject has 2 observations (red and grn)
subject_check <- meta |>
  count(subject_id, name = "n_obs") |>
  filter(n_obs != 2)

#I see 301 participants with data, but EHP paper says 266 with cord blood DNAm data: https://pubmed.ncbi.nlm.nih.gov/35266797/ 
```

#check if all with DNAm have outcomes
```{r}
basc<- read_sas("~/Documents/data/outcomes.sas7bdat") |>
  janitor::clean_names() |>
  filter(visit %in% c("P3", "P4")) |>
  select(1, 2, starts_with("basc"))

#93 participants with DNAm data do not have BASC data
dnam_miss_basc <- meta |>
  distinct(subject_id) |>
  anti_join(
    basc |> distinct(subject_id),
    by = "subject_id"
  )
```

#check participants with exposure during pregnancy
```{r}
pbde_exp <- read_sas("~/Documents/data/pbde.sas7bdat") |>
  janitor::clean_names() |>
  filter(visit %in% c("16W", "26W")) 

ope_exp <- read_sas("~/Documents/data/fr.sas7bdat") |>
  janitor::clean_names() |>
  filter(visit %in% c("16W", "26W", "BIRTH")) 

phth_exp <- read_sas("~/Documents/data/phthbpa.sas7bdat") |>
  janitor::clean_names() |>
  filter(visit %in% c("16W", "26W", "BIRTH"),
         !analyte_code %in% c("BPA")) 

#unique participants with exposures
n_unique_subjects <- bind_rows(
  pbde_exp |> select(subject_id),
  ope_exp  |> select(subject_id),
  phth_exp |> select(subject_id)
) |>
  distinct(subject_id) |>
  nrow()

#exposures with DNAm data
n_subjects_in_meta <- bind_rows(
  pbde_exp |> select(subject_id),
  ope_exp  |> select(subject_id),
  phth_exp |> select(subject_id)
) |>
  distinct(subject_id) |>
  semi_join(
    meta |> distinct(subject_id),
    by = "subject_id"
  ) |>
  nrow()

#exposures with BASC data
n_subjects_in_basc <- bind_rows(
  pbde_exp |> select(subject_id),
  ope_exp  |> select(subject_id),
  phth_exp |> select(subject_id)
) |>
  distinct(subject_id) |>
  semi_join(
    basc |> distinct(subject_id),
    by = "subject_id"
  ) |>
  nrow()

#analytic data with exposures, DNAm , and BASC
n_subjects_analytic <- bind_rows(
  pbde_exp |> select(subject_id),
  ope_exp  |> select(subject_id),
  phth_exp |> select(subject_id)
) |>
  distinct(subject_id) |>
  semi_join(
    meta |> distinct(subject_id),
    by = "subject_id"
  ) |>
  semi_join(
    basc |> distinct(subject_id),
    by = "subject_id"
  ) |>
  nrow()
```

#preprocessing DNAm data
```{r}
# Base folder
file_path <- "~/Documents/k99_home/data/DNAm/dnam_data"  

# List all .idat files recursively (relative paths)
all_files_rel <- list.files(path = file_path, pattern = "\\.idat$", recursive = TRUE)

# Create mapping of file_name -> full path with double slash
file_map <- data.frame(
  file_name = basename(all_files_rel),
  full_path = paste0(file_path, "/", all_files_rel),  # intentional double slash
  stringsAsFactors = FALSE
)

# Build the combined metadata table
cb_dnam_ref <- meta |>
  left_join(file_map, by = "file_name") |>  
  mutate(
    file_type = case_when(
      grepl("Grn", file_name) ~ "file_1",
      grepl("Red", file_name) ~ "file_2"
    ),
    sample_id = str_remove(file_name, "_(Grn|Red)\\.idat$"),
    folder = paste0(vapply(full_path, dirname, character(1)), "/")
  ) |>
  pivot_wider(
    id_cols = c(subject_id, sample_id),
    names_from  = file_type,
    values_from = c(file_name, folder),
    names_glue = "{file_type}_{.value}"
  ) |>
  mutate(
    folder_location = file_1_folder,
    folder_mismatch = file_1_folder != file_2_folder,
    sub_dir = basename(folder_location),
    batch = paste0("Batch", str_extract(sub_dir, "[^-]+$"))
  ) |>
  select(-c(file_1_folder, file_2_folder, 
            file_1_file_name, file_2_file_name, sub_dir,
            folder_mismatch))
```

```{r}
basenames_df_t <- cb_dnam_ref |>
  dplyr::select(sample_id, folder_location) |>
  rename(Sample_ID = sample_id) |>
  arrange(Sample_ID)

# Sort pheno_data_matched_t by basename
pheno_data_matched_t <- cb_dnam_ref |>
  dplyr::select(subject_id, sample_id, batch) |>
  rename(
    participant_id = subject_id,
    basename = sample_id
  ) |>
  arrange(basename)

#check if they match sequence
all(basenames_df_t$Sample_ID == pheno_data_matched_t$basename)
```

#CpGs to drop - Skip this and use updated information provided by Arzu (Feb 2026)
```{r}
#https://pubmed.ncbi.nlm.nih.gov/27330998/ (McCartney 2016)
#https://pubmed.ncbi.nlm.nih.gov/23314698/ (Chen 2013)

chen_crp<- read_csv("~/Documents/HOME_DNAm/drop_cpg/cross_reactive_probes.csv")
chen_failed_probes <- read_csv("~/Documents/HOME_DNAm/drop_cpg/failed_probes.csv")
chen_snp_probes <- read_csv("~/Documents/HOME_DNAm/drop_cpg/snp_probes_minfi.csv")

mcc_polymorphic<- read_delim("~/Documents/HOME_DNAm/drop_cpg/mccarthy_polymorphic.txt",
                             delim = "\t", col_names = TRUE) |> select(c(1)) |>clean_names()
mcc_hybri_targ<- read_delim("~/Documents/HOME_DNAm/drop_cpg/mccarthy_cross_hybri_targ.txt",
                             delim = "\t", col_names = TRUE)
mcc_hybri_non_targ <- read_delim("~/Documents/HOME_DNAm/drop_cpg/mccarthy_corss_hybri_non_targ.txt",
                                 delim = "\t", col_names = TRUE)

all_cpgs_to_drop <- bind_rows(chen_crp, chen_failed_probes, chen_snp_probes,
                              mcc_polymorphic, mcc_hybri_targ, mcc_hybri_non_targ) |>
  distinct()

write_csv(all_cpgs_to_drop, "~/Documents/HOME_DNAm/drop_cpg/all_cpgs_to_drop.csv")
```

#cpgs to drop - Arzu's list (Feb 2026) - includes all filters from Zhou et al. 2017 (https://pubmed.ncbi.nlm.nih.gov/27924034/) 
```{r}
Zhou <- readRDS("~/Documents/k99_home/arzu_cpgs_to_drop/EPIC.hg19.manifest.rds")

Zhou_tbl <- tibble(
  cpg = names(Zhou),
  MASK_general = GenomicRanges::mcols(Zhou)$MASK_general,
  chrm_A = as.character(GenomicRanges::seqnames(Zhou))
) %>%
  filter(
    MASK_general == FALSE,
    !chrm_A %in% c("chrM", "chrX", "chrY"),
    str_starts(cpg, "cg")        
  ) %>%
  select(cpg)

write_csv(Zhou_tbl, "~/Documents/k99_home/arzu_cpgs_to_drop/cpgs_to_keep.csv")
```


#preprocessing steps
#Step-1: Read IDAT files (raw data)
#Step-2: Calculate detection p-values 
#Step-3: Identify failed probes: low-quality probes defined as detection p < 1.0E-7 or number of beads <3 in at least 5% of my total samples
#Step-4: Sample QC: Remove samples with >5% failed probes
#Step-5: Drop failed probes
#Step-6: Drop probes with detection rate <95%
#Step-7: Extremely low intensity of bisulfite conversion probes (less than 3 × standard deviation (SD) of the intensity across samples below the mean intensity)
#Step-8: Noob normalization (background correction and dye bias)
#step-9: BMIQ normalization to correct for probe type bias
#step-10: Batch correction using ComBat
#step-11: Estimate cell proportions using Bakulski et al. Epigenetics 2016 refrence "FlowSorted.CordBlood.450k" to get leucocytes (granulocytes, monocytes, B cells, natural killer [NK] cells, CD8+ T-lymphocytes and CD4+ T-lymphocytes)
#Step-12: Remove sex chromosome probes
#Step-13: Remove all probes listed here: "~/Documents/k99_home/dnam_prep_code/all_cpgs_to_drop.csv"
#Step-14: Compute M-values: M= log2(beta / (1-beta) )


#process
```{r}
process_methylation_data(
  basenames_df = basenames_df_t, 
  pheno_data = pheno_data_matched_t, 
  output_folder = "~/Documents/k99_home/dnam_prep_data", 
  array_type = "EPIC",  
  detection_pval = 1e-6,
  beadcount_threshold = 3,
  sample_cutoff = 0.05,
  probe_detection_rate = 0.95,
  bisulfite_sd_threshold = 3,
  cpgs_to_keep_file = "~/Documents/k99_home/arzu_cpgs_to_drop/cpgs_to_keep.csv",
  compute_mvalues = FALSE,
  apply_bmiq = TRUE,
  apply_combat = TRUE,
  batch_variable = "batch",
  estimate_cell_proportions = TRUE,
  verbose = TRUE
)
```

